Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-02-01 causal-bandit/icml-2015/general.tex

We now consider single variable interventions on general causal graphs over binary variables $Y, X_1,\ldots,X_N$. 
As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of $i$ so the joint distribution over $X$ may be written as
\eq{
\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
}
Throughout this section we assume that the causal structure and conditional distributions of $X$ are known,
which allows us to compute the above quantity for arbitrary $x \in \Omega = \set{0,1}^N$. Given the learner intervens by $do(X_i = j)$
leads to
\eq{
&\Pn{ij}{X = x} = \P{X = x|do(X_i = j)} \\
&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
}
The expected reward by making the intervention $do(X_i = j)$ is
\eq{
\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,.
}
The problem of learning the optimal intervention is made more complicated than the simple case because of 
the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual graph below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\caption{}\label{fig:causalStructure_confounded}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\end{figure}

Another difficulty is that it is no longer sufficient to learn about $\mu_{ij}$ exclusivly from observational data and when $do(X_i = j)$ is chosen.
We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. 
If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to 
explicitly play each of these arms individually to estimate their rewards.

>>>>

We now consider the more general problem where the graph structure is known, but arbitrary.
As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of variable $X_i$ so the joint distribution over $X$ may be written as
\eq{
\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
}
Throughout this section we assume that the causal structure and conditional distributions of $X$ are known,
which allows us to compute the above quantity for arbitrary $x \in \Omega = \set{0,1}^N$. Given the learner intervens by $do(X_i = j)$
leads to
\eq{
&\Pn{ij}{X = x} = \P{X = x|do(X_i = j)} \\
&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
}
The expected reward by making the intervention $do(X_i = j)$ is
\eq{
\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,.
}
The problem of learning the optimal intervention is made more complicated than the simple case because of 
the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual graph below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\caption{}\label{fig:causalStructure_confounded}
\end{figure}

Another difficulty is that it is no longer sufficient to learn about $\mu_{ij}$ exclusivly from observational data and when $do(X_i = j)$ is chosen.
We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.

### 2016-02-01 causal-bandit/icml-2015/introduction.tex

Our main results for this case are: an upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?); an upper bound on the cumulative regret (Theorem~\ref{thm:}) for a specific causal model that has a better dependence on the number of available actions then the best possible algorithm which ignores the causal model; and a matching lower bound (Theorem~\ref{thm:}) for any algorithm which does use the additional causal information.
Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.

>>>>

In \S\ref{sec:parallel} we analyse a specific family of causal bandit problems which we call \emph{parallel bandit} problems.
We  upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?)
\todom{Finish describing new structure}

; an upper bound on the cumulative regret (Theorem~\ref{thm:}) for a specific causal model that has a better dependence on the number of available actions then the best possible algorithm which ignores the causal model; and a matching lower bound (Theorem~\ref{thm:}) for any algorithm which does use the additional causal information.
Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.
